Extracting Function Signature from R Program Traces

### 2020-05-12 paper/main.tex
lists and relationship between list and vectors

To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
That said, we don't' intend to subsume all validation code, a topic we discuss next.

@@

\AT{Will revisit to tighten discussion, need to think about best way to
  phrase this.  Perhaps "dependent type system bad" is enough?}  Including
data dimensions in types would make our type system a dependent type system,
which we believe is too complex to retrofit onto a dynamic data science
language such as R.  The simplest way to include dimensions would be to
explicitly specify the dimensions of an argument (e.g. specifying that a
function should always be called with a double vector of length 10), which
is not particularly useful, and a pattern which \AT{rarely appeared in our
  analysis}.

@@

\AT{List rehash.}  Separate from the vectorized primitives in R are lists:
Whereas vectors are homogeneous and restricted to primitive types, lists can
contain arbitrarily-typed data.  R even builds on lists to create a more
complex data structure known as the {\it data frame}, an essential data
structure to R's functionality \AT{maybe too aggressively worded, but we can
  back this up with data, e.g. how many functions take data frames?}.

\AT{Outline concerns for typing.}  We are faced with several possible
designs for list types.  Chief concerns include the heterogeneity of list
elements, and the presence of a \code{names} attribute, allowing list
elements to be accessed in a struct-like manner.  \AT{Rarely?}, programmers
even write code expecting lists of a particular pre-defined length.  The
next sections highlight some important design decisions for our list types.

\AT{Dig into concerns for typing.}

>>>>

To capture cases like this, we include a \code{class<matrix>} type in our type annotation language.
If \code{rowWeightedMeans} were outfitted with such a type, \contractr can perform the check present in this code automatically.
That said, we don't intend to subsume all argument validation code, discussed next.

@@

Given the discussion of scalars, vectors, and matrices, it might be tempting to consider factoring dimensionality into types.
The primary drawback of this would be rendering our type annotation language into a language of dependent types, which we believe too complex to retrofit onto a dynamic data science language such as R.  
This does mean our types cannot subsume all validation code, such as some later code from the aforementioned \code{rowWeightedMeans} function, shown in Figure~\ref{fig:general-validation-example}.

\begin{lstlisting}
rowWeightedMeans <- function(x, w = NULL, rows = NULL, cols = NULL,
                             na.rm = FALSE, ...) {
  ...
  n <- ncol(x)
  if (length(w) != n) {
    stop("The length of argument 'w' is does not match the number of column in 'x': ", 
    length(w), " != ", n)
  }
  ...
\end{lstlisting}

\caption{General dimensionality check example from the \code{matrixStats} package.}
\label{fig:general-validation-example}
\end{center}
\end{figure}

Here, the length of the weights vector \code{w} is being checked for equality with the number of columns in the matrix \code{x}. 
Ultimately, we decided to leave these kinds of checks up to the programmer.
\AT{I feel like we might want a better argument to not make our type language dependent (better than lol pls no dependent types), but I could be wrong.
Maybe reviewers would be ok with a simple argument.}

@@

Separate from the vectorized primitives in R are lists:
Whereas vectors are homogeneous and restricted to primitive types, lists can contain arbitrarily-typed data.  
Further, lists and vectors have a different syntax for accessing elements, where lists are indexed using \code{[[} and vectors using \code{[}.
R even builds on lists to create a more complex data structure known as the {\it data frame}, an essential data structure to R's functionality.

We are faced with several possible designs for list types.  
Chief concerns include the heterogeneity of list elements, and the presence of a \code{names} attribute, allowing list elements to be accessed in a struct-like manner.  
\AT{Rarely?}, programmers even write code expecting lists of a particular pre-defined length.  
The next sections highlight some important design decisions for our list types.

%
%
\subsubsection{Relationship Between Lists and Vectors}

The ease with which vectors may be transformed into lists is perhaps indicative of a relationship between the two.
In R, all vectors can be converted to lists with a simple call to \code{as.list}, and lists to vectors with \code{unlist}, though care must be taken when unlisting a non-homogeneous list, as coercions will be performed where possible, and the function will simply do nothing if coercion is impossible.
That said, lists and vectors have different access syntax, so vectors cannot stand-in for lists without the \code{as.list} call above.

In sum, we opted not to include a subtyping relationship between vectors and lists, keeping both types separate.
In practice, \AT{how often do \code{T[] | list<T>} types occur?}.
During our evaluation we also found it rare that programmers would attempt to pass list-typed values to vector-typed arguments and vice versa, and we will discuss this further in Section~\ref{sec:evaluation}.
